---
title: "R for Marine Science - Workshop 1"
author: "Darius HUISMAN"
date: "2026-06-03"
output: 
  html_document:
    code_fold: false
    echo: true
    collapse: yes
    df_print: paged
    fig_caption: yes
    fig_height: 6
    fig_width: 10
    highlight: textmate
    theme: default
    toc: true
    toc_float: yes
editor: source
---


```{r setup, include=FALSE}
# Set options, housekeeping
knitr::opts_chunk$set(
	echo = TRUE,
	include = TRUE,
	message = FALSE,
	warning = FALSE)
```

# Introduction

This code contains the work realised for the first workshop of the second module of MB5370.

This workshop focus on the improvement of foundations of data wrangling and plotting for data science. In order to complete this, better workspace setup and repository path calibration was realised. Then tidyverse grammar was used to manipulate data to provide with descriptive data summaries and adequate figure using ggplot2.

```{r housekeeping, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
```

```{r packages, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse) 
library(readxl)
library(here)
library(palmerpenguins)
```

```{r source-script1, message=FALSE, warning=FALSE, include=FALSE}
source("Workshop1.R")
```

# 1. Ingesting data into R
```{r import_data, include=TRUE}
# Practice Import A: Loading a standard comma-separated plain text file
benthic_cover <- read_csv(here::here("data/workshop1/reef_cover_log.csv"))

# Practice Import B: Parsing a tab-separated telemetry instrument array string
acoustic_stream <- read_tsv(here::here("data/workshop1/acoustic_telemetry_stream.txt"))

# Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
fisheries_annual <- read_excel(here::here("data/workshop1/fish_catch_data.xlsx"), sheet = "Commercial_2026")

# Read in mangrove_data
mangrove_data <- read_csv(file = here::here("data/workshop1/mangrove_survey_raw.csv"))

# Use args within read_csv to skip headers and declare missing flags
mangrove_data <- read_csv(
  here::here("data/workshop1/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA
```


# 2. Data frame architectures: Tibbles versus legacy tables
```{r tibbles_and_table}
# Force a modern tibble to degrade into a legacy base R data frame structure
benthic_cover_df <- as.data.frame(benthic_cover)

# Print the old-style dataframe structure to view
print(benthic_cover_df)
# And compare with tibble alternative
print(benthic_cover)
```

# 3. Wrangling out ecological signals using Palmer Penguins
```{r palmerpenguins_data}
# Load the package data into active memory
data("penguins")

# Examine the structure of the dataset - always do this when loading a new dataset!
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version

# Generate an exploratory summary matrix
summary(penguins)
```

# 4 Foundational grammar: Slicing, filtering, sorting, and transforming
## 4.1 Isolating attributes with select()
```{r palmerpenguins_select}
# Vertically slice specific morphometric variables by explicit name
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)

# Retain a continuous block of attributes using the colon operator
spatial_block <- select(penguins, species:island)

# Discard logistics tracking attributes while preserving everything else using the minus sign
clean_scientific_fields <- select(penguins, -year)
```

## 4.2 Sifting rows with filter()
```{r palmerpenguins_filter}
# Isolate observations belonging to a single categorical target group
adelie_cohort <- filter(penguins, species == "Adelie")

# Sift out individuals using continuous numerical boundary thresholds
# Preserves only large penguins whose mass exceeds 4500 grams
heavy_penguins <- filter(penguins, body_mass_g > 4500)

# Combine multiple conditional parameters across separate attributes
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")

# Sift records matching multiple targeting flags within an explicit set
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

## 4.3 Ordering sequences with arrange()
```{r palmerpenguins_arrange}
# Sort penguins by ascending body mass (Default setting: Smallest mass first)
lightest_first <- arrange(penguins, body_mass_g)

# Sort penguins in descending sequence using the desc() layout wrapper
heaviest_first <- arrange(penguins, desc(body_mass_g))

# Execute nested sorting criteria: Group by species, then sort by descending bill length
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

## 4.4 Introducing the Pipe (|>)
```{r palmerpenguins_bill_ratio}
# use a pipe to calculate bill ratio of a specific species
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```

## 4.5 Computing new attributes with mutate()
```{r palmerpenguins_body_mass}
# Calculate a new morphological ratio in our environment
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

# View your newly engineered variables appended to the far-right columns
glimpse(penguin_ratios)
```

# 5. Data aggregation and ecological summarisation
```{r palmerpenguins_group_&_summary}
# Grouping our active memory penguins by species
grouped_penguins <- group_by(penguins, species)

# Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)

# Collapsing the buckets into explicit summary metrics
species_mass_summary <- summarise(grouped_penguins,
                                  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)

# Overcoming the missing value trap using na.rm = TRUE
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)
```


# 6. Integrating data grammar with visual diagnostics in qmd

## 7.1 Wrangling and Plotting in Parallel

```{r Wrangling and Plotting in Parallel}
#summary table
penguins  |>
  group_by(species, island) |>
  summarise(mean_bodymass = mean(body_mass_g, na.rm = TRUE)) 
penguins|> 
  ggplot(aes(x = species, y = body_mass_g, fill =island)) + 
  geom_boxplot()  
```

## 7.2 Piping Directly to Visualisation
```{r Piping Directly to Visualisation}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```

Pause for thought: Think about what this figure is telling you about the biology of these species of penguins on these different islands. What do you think a statistical test would tell you here?
  The bodymass correlates with the species and not the island.




```{r challenge (SE instead of SD)}
# Pipe directly from aggregation to plotting with error bars
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    se_mass = (sd(body_mass_g, na.rm = TRUE)/sqrt(length(body_mass_g))),
    n = n(),
    .groups = "drop"
  ) |>
  ggplot(aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - se_mass, 
                    ymax = mean_mass + se_mass), 
                width = 0.2) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard errors",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```


# 8. Saving, exporting, and version milestones

```{r Create output directories if they do not exist}
# Create output directories if they do not exist:
if (!dir.exists("outputs/figures")) dir.create("outputs/figures") # folder for figs
if (!dir.exists("outputs/tables")) dir.create("outputs/tables") # folder for tables
if (!dir.exists("Rdata")) dir.create("Rdata") # folder for Rdata objects

```

```{r export tables}
# 1. Exporting our collapsed summary table as a universal flat text file
write_csv(biological_signal, "outputs/tables/penguin_species_mass_summary.csv")

# 2. Saving our penguins_final table as a native R binary file
saveRDS(penguins_final, "outputs/tables/penguins_final.rds")

```

```{r exprot figures}
ggsave("outputs/figures/mass_compare_plot.png", 
       plot = mass_compare_plot, 
       width = 120, height = 120, 
       units = "mm", dpi = 300)
```

